ODGI: scalable tools for working with pangenome graphs

### 2021-05-29 manuscript/main.tex
added non-breaking space before the references

    A pangenome models the full set of genomic elements in a given species or clade \citep{32453966}. Indeed, these
    data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based
    approaches which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
    pangenomes involves the \textit{sequence graphs} \citep{2488477}, where similar orthologous regions between genomes
    collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
    sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph. A
    \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{30125266}.
    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

    Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate. This amount of
    data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to
    represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full
    genetic variation across a population, but graph-based data structures present computational overheads: in addition
    to the sequences, they must also represent the graph topology. Several genome data structures were developed to
    address these challenges \citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
    graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform
    efficiently on the complex graphs generated from whole-genome alignment of vertebrates. Highly repetitive regions,
    such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs which present paths with
    very high-depth nodes. The node depth is defined as the number of times in which the node is crossed by all the
    paths embedded in the graph. High depth increases the complexity of the operations performed on the paths and the
    nodes they cross.

    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
    loads graphs into the succinct dynamic HandleGraph model \citep{33040146} and applies methods based on the
    HandleGraph API. Several algorithms are shared with the VG toolkit (VG) \citep{30125266}, the state-of-the-art
    toolkit for working with pangenome graphs. However, although similar in scope to VG, ODGI importantly differs in
    its design to support highly complex pangenome graphs. The majority of the ODGI tools are implemented in an
    index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
    implementations would require building indexes also for single or little operations on the graphs. Moreover,
    thanks to its efficient path representation, ODGI tools can build the internal graph representation (the ODGI
    format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
    efficiently scaling up to hundreds of large vertebrate genomes.

@@

        ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
        applies concepts first introduced in the dynamic version of the Graph BWT \citep{31406990}, a scalable
        implementation of the graph extension of the positional Burrows-Wheeler transform \citep{28702075}. Specifically,
        we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
        The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the
        node, its edges in both directions (forward and backward), and a vector of \textit{path steps} that describes
        how the embedded paths walk across the node. Each step records the path identifier and the previous and next
        path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the
        \textit{N} neighbors of the node into the range \textit{1..N}. To further save space, \textit{node deltas},
        rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the
        graph vector (i.e., the difference between the node offsets). The path step vector, of length \textit{n}, is
        stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
        \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node structure includes a
        byte-width mutex lock.

@@

    \begin{figure}[!tpb]%figure1
        \fboxsep=0pt\colorbox{gray}{
            \begin{minipage}[t]{235pt}
                \vbox to 100pt{
                    \vfill\hbox to
                    235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}
                    \vfill
                }
            \end{minipage}}
%\centerline{\includegraphics{fig01.eps}}
        \caption{
            Visualizing the Major Histocompatibility Complex (MHC) \textit{locus} pangenome. A) odgi viz 1D
            visualization of 8 haploid, phased human genome assemblies plus the chm13 and GRCh38 reference genomes.
            Graph nodes’ are arranged from left to right. The colored bars represent the linearized renderings of the
            embedded paths versus the nodes in a binary matrix. The black lines under the paths are the links, which
            represent the graph topology. B) Bandage \citep{26099265} representation of the graph topology of a
            consensus MHC pangenome graph representing only variations bigger than 100 bps. Gene labels were applied
            exploiting \texttt{odgi position} to liftover gene coordinates from the full graph into the consensus
            graph.
        }\label{fig:01}
    \end{figure}

>>>>

    A pangenome models the full set of genomic elements in a given species or clade~\citep{32453966}. Indeed, these
    data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based
    approaches which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
    pangenomes involves the \textit{sequence graphs}~\citep{2488477}, where similar orthologous regions between genomes
    collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
    sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph. A
    \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{30125266}.
    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

    Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate. This amount of
    data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to
    represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full
    genetic variation across a population, but graph-based data structures present computational overheads: in addition
    to the sequences, they must also represent the graph topology. Several genome data structures were developed to
    address these challenges~\citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
    graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform
    efficiently on the complex graphs generated from whole-genome alignment of vertebrates. Highly repetitive regions,
    such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs which present paths with
    very high-depth nodes. The node depth is defined as the number of times in which the node is crossed by all the
    paths embedded in the graph. High depth increases the complexity of the operations performed on the paths and the
    nodes they cross.

    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
    toolkit for working with pangenome graphs. However, although similar in scope to VG, ODGI importantly differs in
    its design to support highly complex pangenome graphs. The majority of the ODGI tools are implemented in an
    index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
    implementations would require building indexes also for single or little operations on the graphs. Moreover,
    thanks to its efficient path representation, ODGI tools can build the internal graph representation (the ODGI
    format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
    efficiently scaling up to hundreds of large vertebrate genomes.

@@

        ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
        applies concepts first introduced in the dynamic version of the Graph BWT~\citep{31406990}, a scalable
        implementation of the graph extension of the positional Burrows-Wheeler transform~\citep{28702075}. Specifically,
        we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
        The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the
        node, its edges in both directions (forward and backward), and a vector of \textit{path steps} that describes
        how the embedded paths walk across the node. Each step records the path identifier and the previous and next
        path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the
        \textit{N} neighbors of the node into the range \textit{1..N}. To further save space, \textit{node deltas},
        rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the
        graph vector (i.e., the difference between the node offsets). The path step vector, of length \textit{n}, is
        stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
        \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node structure includes a
        byte-width mutex lock.

@@

    \begin{figure}[!tpb]%figure1
        \fboxsep=0pt\colorbox{gray}{
            \begin{minipage}[t]{235pt}
                \vbox to 100pt{
                    \vfill\hbox to
                    235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}
                    \vfill
                }
            \end{minipage}}
%\centerline{\includegraphics{fig01.eps}}
        \caption{
            Visualizing the Major Histocompatibility Complex (MHC) \textit{locus} pangenome. A) odgi viz 1D
            visualization of 8 haploid, phased human genome assemblies plus the chm13 and GRCh38 reference genomes.
            Graph nodes’ are arranged from left to right. The colored bars represent the linearized renderings of the
            embedded paths versus the nodes in a binary matrix. The black lines under the paths are the links, which
            represent the graph topology. B) Bandage~\citep{26099265} representation of the graph topology of a
            consensus MHC pangenome graph representing only variations bigger than 100 bps. Gene labels were applied
            exploiting \texttt{odgi position} to liftover gene coordinates from the full graph into the consensus
            graph.
        }\label{fig:01}
    \end{figure}
